---
title: "Problem Set 2"
author: "Brendan Case"
documentclass: article
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Exponential growth in cancer

The growth of the cancer cell follows the formula for exponential growth:

$$N_{t+1} = 2 N_{t},$$

where $t$ is a timestep of 109 days. If $N_0 = 1$, then the above equation simplifies to $N_{t+1} = 2^{t}$ and we must solve

$$10^{12} = 2^{t_{crit}},$$

or

$$t_{crit} = \log_2(10^{12}),$$

which is `r log(10^(12), 2)`. Since a time-step is 109 days, the number of days until the critical value is roughly $109\cdot 40=`r 109 * 40`$. 

## 2. Sterile insect release

Let

$$F(N) = b \frac{N}{N + M} - d.$$

Throughout this section, I will suppose $b, d$, and $M$ are non-negative. Further, population levels of $N$ below zero we consider to be simply extinct (stuck at 0).

### *Equilibrium points and stability*

I will write the equilibrium points as $N^*$, i.e., we seek solutions to the equation
$F(N^*) = 0$. Solving this for $N^*$, we see that

\begin{equation*}
  N^* = \begin{cases}
    0, & \text{when $b = d$}\\
    \frac{d}{b - d}M & \text{otherwise.}
  \end{cases}
\end{equation*}

Note that these cases do not indicate there are two equilibrium points.

Further, since applying the quotient rule for $f(N) = N$ and $g(N) = N + M$ quickly gives $\frac{\text{d}F}{\text{d}N} = b\frac{M}{(N+M)^2} > 0$, and we know from class that $N^*$ is stable if and only if $\frac{\text{d}F}{\text{d}N} < 0$, $N^*$ is unstable for any parameter setting.

Finally, for completeness I should point out that by assuming that if $N$ has crashed to less than or equal to 0, $N$ will stay at 0, I have effectively insisted that $N*=0$ is a stable equilibrium point.

### *Critical threshold for population extinction*

Since from the previous section we know values of $N$ above $N^*$ will continue to grow away from $N^*$, to demonstrate that $N^*$ is the critical threshold it remains to show values of $N$ below $N^*$ will be attracted to 0. However, this immediate after noting $N/(N + M)$, and hence $F$, is strictly monotone increasing:

```{r}
X = 1:100
Y1 = X/(X + 1)
Y2 = X/(X + 500)
plot(X, Y1, ylim=c(0, 1), ylab="N/(N+M)", xlab="N")
points(X, Y2)
```

This can be proven after noting the first derivative of $N/(N+M)$ is clearly non-negative.

Since $F$ is strictly monotone, this means

$$F(N) < F(N^*) = 0$$

for $0 \leq N < N^*$, so $N$ will always decrease toward 0 in this case.

In summary, this model leads to a strong Allee effect:

```{r}
Y3 = 2.4 * X/(X + 50) - .5
plot(X, Y3, xlab="N", ylab="F")
abline(h=0)
```

As can be seen, $F$ scales with population size $N$ such that if $N$ is below the critical value, the population will crash, but past this value, growth will increase further as $N$ increases. 

<!-- In particular, when $N$ is rather small (close to 0), $N$ must be increasing since $N^*=0$ is not stable. Similarly, if $N < \frac{d}{b - d}M$ then the population will be decreasing, and increasing when $N > \frac{d}{b - d}M$. Therefore, $\frac{d}{b - d}M$ is the critical threshold for determining extinction. The population of female mosquitoes is subject to a strong Allee effect in this model. -->

## 3. Sterile insect release with competition

```{r, message=F}
require(deSolve)
times <- seq(0,500,by=0.1)

# Parameters
params <- c(b = 2.4,c=0.02,M=50,d=0.2)

# Model
sterile_insect <- function(t,state,parameters){
  with(as.list(c(state,parameters)),{
    if (N <= 0) {
      dN = 0
    }
    else {
      dN <- b*N/(N+M) - d - c*N
    }
    return(list(c(dN)))})
}
```

To answer this question, I will first vary the initial $N_0$ and then plot the last 10 values of $N$ after 5,000 timesteps. 

```{r}
plot(1, type="n", xlab="initial value N0", ylab="final values N", xlim=c(0, 200), ylim=c(0, 55))
for (n in seq(6, 200, by=1)) {
  out <- ode(y = c(N=n), times=times, func=sterile_insect, parms=params)
  points(x=rep(n, 10), y=tail(out[,2], 10))
}
```

This suggests there are three equilibrium points at $N^*=0, 10, 50$ for these particular parameter settings. Further, since for $N_0 < 10$ and $N_0 > 50$ the model was attracted to the points $0$ and $50$ respectively, the points $0$ and $50$ are stable, while $10$ is not.

```{r}
diff_eq = function(N) {
  return(params[1] * (N/(N + params[3])) - params[4] - params[2] * N)
}

X = 0:100
Y = diff_eq(0:100)
plot(X, Y,
     xlim=c(0, 100), 
     xlab="N",
     ylab="dN",
     pch=ifelse(X %in% c(0, 50), 19, 1), 
     cex=1.6)
```


